\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-28 tex/paper.tex
save

In this note, we develop a likelihood-free inference algorithm for the point
estimation of the parameters of a forward non-differentiable
generative model. We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an arbitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

\subsection{Toy problems}

As illustrative experiments, we evaluate adversarial variational optimization on
well-known distributions for continuous and discrete data. We artificially
consider these distributions as parameterized black box simulators, from which
we can only generate data.

{\it Continuous data.}
We first consider inference for
a univariate Gaussian with unknown mean and
unknown variance. The observed data is sampled from a univariate Gaussian with
mean $\mu^* = 1$ and standard deviation ${\sigma^*} = 0.5$. Algorithm~\ref{alg:avo} is run
for 300 epochs with mini-batches of size $M=64$ and the following configuration.
For the critic, we use a 3-layer MLP with 10 hidden nodes per layer and ReLU
activations. At each epoch, Adam is run for $n_\text{critic}=100$ iterations
with a step size $\alpha=0.01$, decay rates $\beta_1=\beta_2=0.5$ and its inner
first and second moment vectors reset at each epoch in order to avoid building
momentum in staled directions.  For estimating $\theta^* = (\mu^*, \sigma^*)$, we use
factored univariate Gaussian proposal distributions $q(\mu|\psi)$ and $q(\log
\sigma|\psi)$ initialized with zero mean and unit variance. At each epoch,
parameters $\psi$ are updated by making one Adam step, with $\alpha=0.01$ and
$\beta_1=\beta_2=0.1$. Penalty coefficients are set to $\lambda=0.025$ and
$\gamma=5$.

\begin{figure}
    \centering
    \vspace{-0.8cm}
    \includegraphics[width=0.48\textwidth]{figures/gaussian-gamma=5.pdf}
    \vspace{-1cm}
    \caption{Toy problem on continuous data: Gaussian with unknown mean and unknown variance.
             (Top) Model distribution $p(\mathbf{x}|\psi)$ after adversarial variational optimization along with the proposal distributions $q(\mu|\psi)$ and $q(\log
             \sigma|\psi)$ over the generator parameters.
             (Bottom) Evolution of the variational upper bound $U_d$ on the loss ${\cal L}_d$ of the critic. }\label{fig:gaussian}
\end{figure}
\begin{figure}
    \centering
    \vspace{-0.5cm}
    \includegraphics[width=0.48\textwidth]{figures/poisson-gamma=5.pdf}
    \vspace{-1cm}
    \caption{Toy problem on discrete data: Poisson with unknown scale.
             This example illustrates adversarial variational optimization on discrete non-differentiable generators.}\label{fig:poisson}
\end{figure}

The upper plot of Figure~\ref{fig:gaussian} illustrates the resulting model
after the optimization procedure. As theoretically expected from adversarial
training, the model density $p(\mathbf{x}|\psi)$ closely matches the true
density $p_r(\mathbf{x})$, as shown by the aligned histograms in blue and
orange. Under the effect of the entropy penalty $\gamma H(q(\theta|\psi))$, we
also observe that the proposal distributions have correctly concentrated their
mass around $\mu^*=1$ and $\log \sigma^*=-0.69$. The bottom plot of
Figure~\ref{fig:gaussian} shows an empirical estimate of $-U_d$ with respect to
the epoch. We observe that it quickly falls towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)}
[d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$ and that the critic cannot distinguish
between true and model data.

{\it Discrete data.}
Similarly, we evaluate adversarial variational optimization on discrete data,
for the inference of the scale parameter $\lambda$ of a Poisson distribution.
The observed data is sampled from a univariate Poisson with scale $\lambda^*=7$.
We use a univariate Gaussian proposal distribution $q(\log \lambda|\psi)$
initialized with a mean at $\log 5$ and unit variance. The optimization setup is
otherwise the same as for the continuous case. Despite the discreteness and the non-differentiability
of the underlying generator, Figure~\ref{fig:poisson} shows that inference with
adversarial variational optimization works. The upper plot in the figure shows
that $p(\mathbf{x}|\psi)$ closely matches $p_r(\mathbf{x})$, with the mass of
the proposal distribution concentrated at $\log 7 = 1.94$.

\subsection{Galton box}

>>>>

In this note, we develop a likelihood-free inference algorithm for the point
estimation of the parameters of a forward non-differentiable
generative model. We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an arbitrarily tight family of models whose
joint collection of generated samples matches the observed data.
\glnote{add note on discrete data}

@@

\subsection{Univariate discrete data}

\glnote{contour plots of q(theta|psi) + histograms + learning curves}
\glnote{contours maybe with $\gamma=0$ and $\gamma>0$ contours}
\glnote{insist that we dont learn the simulator}

As a first illustrative experiment, we evaluate adversarial variational
optimization on a Poisson distribution with unknown scale parameter $\lambda$.
We artificially consider the distribution as a parameterized black box
simulator, from which we can only generate (discrete) data.

The observed data is sampled from a Poisson with scale $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 300 epochs with mini-batches of size $M=64$
and the following configuration. For the critic, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=100$ iterations with a step size $\alpha=0.01$, decay rates
$\beta_1=\beta_2=0.5$ and its inner first and second moment vectors reset at
each epoch in order to avoid building momentum in staled directions.  For
estimating $\lambda^*$, we use a univariate Gaussian proposal distribution
$q(\log \lambda|\psi)$ initialized with a mean at $\log(5)$ and unit variance. At
each epoch, parameters $\psi$ are updated by making one Adam step, with
$\alpha=0.01$ and $\beta_1=\beta_2=0.1$. Penalty coefficients are set to
$\lambda=0.025$ and $\gamma=5$.

\begin{figure}
    \centering
    \vspace{-0.5cm}
    \includegraphics[width=0.48\textwidth]{figures/poisson-gamma=5.pdf}
    \vspace{-1cm}
    \caption{Toy problem on discrete data: Poisson with unknown scale.
             This example illustrates adversarial variational optimization for a discrete non-differentiable generator.
             (Top) Model distribution $p(\mathbf{x}|\psi)$ after adversarial variational optimization along with the proposal distribution $q(\lambda|\psi)$.
             (Bottom) Evolution of the variational upper bound $U_d$ on the loss ${\cal L}_d$ of the critic.
             }\label{fig:poisson}
\end{figure}

The upper plot of Figure~\ref{fig:poisson} illustrates the resulting model
after the optimization procedure. As theoretically expected from adversarial
training, the model density $p(\mathbf{x}|\psi)$ closely matches the true
density $p_r(\mathbf{x})$, as shown by the aligned histograms in blue and
orange. Under the effect of the entropy penalty $\gamma H(q(\theta|\psi))$, we
also observe that the proposal distribution has correctly concentrated its
mass around $\log(7) = 1.94$. The bottom plot of
Figure~\ref{fig:poisson} shows an empirical estimate of $-U_d$ with respect to
the epoch. We observe that it quickly falls towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)}
[d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$ and that the critic cannot distinguish
between true and model data. Despite the discreteness and the non-differentiability
of the underlying generator, this confirms that inference with
adversarial variational optimization works.

\subsection{Multidimensional continuous data}

\glnote{use 5d example from carl}
\glnote{contour plots}

\subsection{Weinberg}
